---
title: "TCR-seq"
output: html_notebook
author: "Hernan Lorenzi"
date: 04-07-2023
---

```{r}
library(immunarch)
library(cowplot)
library(R.utils)
```

# Build working dir structure
```{r}
dir.create(path = "./Plots", showWarnings = F)
```


```{r}
immdata <- repLoad(.path = "./TCR_immunarch/")
```

# Select LN samples
```{r}
LN_immdata <- immdata
LN_sample_ids <- filter(immdata$meta, Tissue == "LN")$Sample
LN_immdata$data <- immdata$data[LN_sample_ids]
LN_immdata$meta <- filter(immdata$meta, Tissue == "LN")

```

# Remove all NA rows from sample 1798
```{r}
#LN_immdata_filter$data$TCR_LN_1798_Het[!is.na(LN_immdata_filter$data$TCR_LN_1798_Het$Clones),]
```


# Filter out those clones with only one A or B chain predicted to ensure that cells come from the same clone
```{r}
LN_immdata_filter <- LN_immdata
for (sample in LN_sample_ids){
  LN_immdata_filter$data[[sample]] <- LN_immdata$data[[sample]][ str_detect(string = LN_immdata$data[[sample]]$chain, pattern = "TRA;TRB"),]
}
```

# Plot number of unique clonotypes per genotype
```{r}
exp_vol <- repExplore(LN_immdata_filter$data, .method = "volume")
p1 <- vis(exp_vol, .by = c("Genotype"), .meta = LN_immdata_filter$meta)
ggsave2(filename = "unique_clonotype_per_genotype.pdf", plot = p1, path = "./Plots/")
print(p1)
```

```{r}
exp_len <- repExplore(LN_immdata_filter$data, .method = "len", .col = "aa")
exp_cnt <- repExplore(LN_immdata_filter$data, .method = "count")
exp_vol <- repExplore(LN_immdata_filter$data, .method = "volume")

p1 <- vis(exp_len)
p2 <- vis(exp_cnt)
p3 <- vis(exp_vol, .meta = LN_immdata_filter$meta)

ggsave2(filename = "distr_CDR3_length.pdf", plot = p1, path = "./Plots/")
ggsave2(filename = "distr_clonotype_abund.pdf", plot = p2, path = "./Plots/")
ggsave2(filename = "Numb_unique_clonotypes.pdf", plot = p3, path = "./Plots/")

print(p1)
print(p2 + p3)

```


```{r}
p4 <- vis(exp_len, .by = "Genotype", .meta = LN_immdata_filter$meta)
ggsave2(filename = "distr_CDR3_length_by_Genotype.pdf", plot = p4, path = "./Plots/", height = 8, width = 18)
print(p4)
```





# Clonal proportions: percentage of clonotypes required to occupy a specified percent of the total immune repertoire.
```{r}
imm_pr <- repClonality(LN_immdata_filter$data, .method = "clonal.prop",.perc = 5)
imm_pr
```

# Estimate relative abundance for the groups of top clonotypes in repertoire, e.g., ten most abundant clonotypes. 
```{r}
imm_top <- repClonality(LN_immdata_filter$data, .method = "top", .head = c(10, 100, 1000, 3000, 10000))
print(imm_top)
top.p <- vis(imm_top, .meta = LN_immdata_filter$meta) + vis(imm_top, .by = c("Genotype"), .meta = LN_immdata_filter$meta)
ggsave2(filename = "top_clonal_proportion_by_Genotype.pdf", plot = top.p, path = "./Plots/", height = 8, width = 11)
print(top.p)
```

# Estimate relative abundance for the groups of rare clonotypes with low counts 
```{r}
imm_rare <- repClonality(LN_immdata_filter$data, .method = "rare")
print(imm_rare)
rare.p <- my.vis.immunr_rare_prop(imm_rare, .meta = LN_immdata_filter$meta) + vis(imm_rare, .by = "Genotype", .meta = LN_immdata_filter$meta)
ggsave2(filename = "rare_clonal_proportion_by_Genotype.pdf", plot = rare.p, path = "./Plots/", height = 8, width = 11)
print(rare.p)
```

# Analyse relative abundance (also known as clonal space homeostasis), which is defined as the proportion of repertoire occupied by clonal groups with specific abundances:
```{r}
imm_hom <- repClonality(LN_immdata_filter$data,
  .method = "homeo",
  .clone.types = c(Small = .0001, Medium = .001, Large = .01, Hyperexpanded = 1)
)
print(imm_hom)

hom.p <- vis(imm_hom, .meta = LN_immdata_filter$meta) + vis(imm_hom, .by = c("Genotype"), .meta = LN_immdata_filter$meta)
ggsave2(filename = "rel_abundance_clonotypes.pdf", plot = hom.p, path = "./Plots/", height = 8, width = 11)
print(hom.p)
```

```{r}
overlap.p <- repOverlap(LN_immdata_filter$data) %>% vis()  # Build the heatmap of public clonotypes shared between repertoires
ggsave2(filename = "clonotype_overlap.pdf", plot = overlap.p, path = "./Plots/", height = 8, width = 11)
print(overlap.p)
```

```{r}
LN_immdata_filter_not_single_counts <- LN_immdata_filter
for (sample in LN_sample_ids){
  # Remove rows with counts = 1
  LN_immdata_filter_not_single_counts$data[[sample]] <- filter(LN_immdata_filter$data[[sample]], Clones > 1)
  # Visualise the V-gene distribution across samples
  usage.p <- geneUsage(LN_immdata_filter_not_single_counts$data[[sample]]) %>% vis()  
  ggsave2(filename = paste0("clonotype_usage_",sample,".pdf"), plot = usage.p, path = "./Plots/", height = 8, width = 11)
  print(usage.p)
}
```

```{r}
diversity.p <- repDiversity(LN_immdata_filter$data) %>% vis(.by = c("Genotype"), .meta = LN_immdata_filter$meta)  # Visualise the Chao1 diversity of repertoires, grouped by the patient status
ggsave2(filename = "clonotype_diversity.pdf", plot = diversity.p, path = "./Plots/", height = 7, width = 7)
print(diversity.p)
```

# Load Seurat object and all necessary libraries
```{r}
# Libraries
library(Seurat)

# Load Seurat object
ln_merged_seurat_filtered <- readRDS(file = "./GEX_LN_seurat_analysis_03272023.rds")
```

# Aux funtions to filter TCR cells based on Seurat cluster ID
```{r}
# Function that extracts seurat cluster info from seurate obj metadata
get_seurat_cluster_barcodes <- function(cluster_number, seurat_obj){
  .s_barcodes <- rownames(seurat_obj@meta.data)[seurat_obj@meta.data$seurat_clusters == cluster_number]
  
  # Parse s_barcodes
  .s_barcodes <- separate(data = tibble(.s_barcodes), 
           into = c("tissue", "mouse", "genotype", "barcode"), 
           sep="_", 
           col = ".s_barcodes") %>% 
              mutate(sample = paste('TCR',tissue,mouse,genotype, sep = "_"))
  # Add cluster ID
  .s_barcodes$seurat_clusters <- cluster_number
  .s_barcodes$cell_type <- subset(seurat_obj@meta.data, seurat_clusters == cluster_number)[1,'customclassif']
  return(.s_barcodes)
}

# Function to filter TCR files based on seurat cluster number
# Filtered csv files should be stored in directory called **S_CLUSTER_#** within 
# working dir. Should also include the metadata fil=e for compatibility with immunarch
# Imput files should be named as **sample_name.csv**


filter_tcr_table <- function(s_barcodes, tcr_source_dir){
  .sc_number = s_barcodes$seurat_clusters[1]
  # create directory to store seurat-cluster-specific TCR data
  .sc_dirname = paste0("./TCR_BY_CLUSTER/CLUSTER_",.sc_number)
  dir.create(path = .sc_dirname , showWarnings = F, recursive = T)
  
  
  samples_with_data <- c()
  for (s in unique(s_barcodes$sample)){
    .infilename = paste0(tcr_source_dir,"/",s,".csv")
    .outfilename = paste0(.sc_dirname,"/",s,".csv")
    .tcr_data <- read_csv(file = .infilename, col_names = T)
    .tcr_data_filter <- filter(.tcr_data, barcode %in% s_barcodes$barcode)
    
    # Save sample only if contains any data for current cluster
    if (nrow(.tcr_data_filter) > 0){
      write_csv(x = .tcr_data_filter, file = .outfilename, quote = "none")
      samples_with_data <- c(samples_with_data, s)
    }
  }  
  
  # Load original metadata and add column with Seurat cluster ID
  .inmeta <- read_delim(file = paste0(tcr_source_dir,"/metadata.txt"), col_names = T, delim = "\t")
  .inmeta$Seurat_clusters <- .sc_number
  
  # Remove empty samples from metadata table
  .inmeta <- filter(.inmeta, Sample %in% samples_with_data)
  write_delim(x = .inmeta, file = paste0(.sc_dirname,"/metadata.txt"), 
              delim = "\t", quote = "none")
}
```

# Generate seurat cluster-specific TCR files for further analysis
```{r}
seurat_clusters <- unique(ln_merged_seurat_filtered@meta.data$seurat_clusters)
for (sc_number in seurat_clusters){
  # Retreive seurat cluster barcodes from seurat object
  s_cluster_number = sc_number
  s_barcodes <- get_seurat_cluster_barcodes(cluster_number = s_cluster_number, 
                                            seurat_obj = ln_merged_seurat_filtered)
  
  # Generate seurat cluster-specific TCR files
  filter_tcr_table(s_barcodes = s_barcodes, tcr_source_dir = "./TCR_immunarch_LN")
}
```

# Aux function to run entire immunarch analysis for one cluster
```{r}
# Function to run entire immunarch analysis for one cluster
run_immunarch <- function(imput_dir){
  # Prepare working dir
  .plots_dir = paste0(imput_dir,"/Plots")
  dir.create(path = .plots_dir, showWarnings = F, recursive = T)
  
  # create a link to metadata.txt within Plots to avoid errors loading the metadata with the package immunarch
  createLink(link=paste0(.plots_dir,'/metadata.txt'), target=paste0(imput_dir,'/metadata.txt'))
  
  # Check if metadata contains more than one genotype, otherwise skip this cluster
  .inmeta <- read_delim(file = paste0(imput_dir,'/metadata.txt'), col_names = T, delim = "\t")
  if(length(unique(.inmeta$Genotype)) < 2){
    return()
  }
  
  
  # Load data as an immunarch object
  .immdata <- repLoad(.path = imput_dir)
  
  # Filter out those clones with only one A or B chain predicted to ensure that cells come from the same clone
  .sample_ids <- .immdata$Sample
  .immdata_filter <- .immdata
  for (sample in .sample_ids){
    .immdata_filter$data[[sample]] <- .immdata$data[[sample]][ str_detect(string = .immdata$data[[sample]]$chain, pattern = "TRA;TRB"),]
  }
  
  # Plot number of unique clonotypes per genotype
  .exp_vol <- repExplore(.immdata_filter$data, .method = "volume")
  p1 <- vis(.exp_vol, .by = c("Genotype"), .meta = .immdata_filter$meta)
  ggsave2(filename = "unique_clonotype_per_genotype.pdf", plot = p1, path = .plots_dir)
  
  # Exploratory analysis
  .exp_len <- repExplore(.immdata_filter$data, .method = "len", .col = "aa")
  .exp_cnt <- repExplore(.immdata_filter$data, .method = "count")
  .exp_vol <- repExplore(.immdata_filter$data, .method = "volume")
  
  p1 <- vis(.exp_len)
  p2 <- vis(.exp_cnt)
  p3 <- vis(.exp_vol, .meta = .immdata_filter$meta)
  
  ggsave2(filename = "distr_CDR3_length.pdf", plot = p1, path = .plots_dir)
  ggsave2(filename = "distr_clonotype_abund.pdf", plot = p2, path = .plots_dir)
  ggsave2(filename = "Numb_unique_clonotypes.pdf", plot = p3, path = .plots_dir)
  
  #
  
  p4 <- vis(exp_len, .by = "Genotype", .meta = .immdata_filter$meta)
  ggsave2(filename = "distr_CDR3_length_by_Genotype.pdf", plot = p4, 
          path = .plots_dir, height = 8, width = 18)
  
  # Clonal proportions: percentage of clonotypes required to occupy a specified percent of the total immune repertoire.
  
  .imm_pr <- repClonality(.immdata_filter$data, .method = "clonal.prop",.perc = 5)
  
  # Estimate relative abundance for the groups of top clonotypes in repertoire, e.g., ten most abundant clonotypes.
  
  .imm_top <- repClonality(.immdata_filter$data, .method = "top", .head = c(10, 100, 1000, 3000, 10000))
  top.p <- vis(.imm_top, .meta = .immdata_filter$meta) + 
           vis(.imm_top, .by = c("Genotype"), .meta = .immdata_filter$meta)
  ggsave2(filename = "top_clonal_proportion_by_Genotype.pdf", 
          plot = top.p, path = .plots_dir, height = 8, width = 11)
  
  # Estimate relative abundance for the groups of rare clonotypes with low counts 
  
  .imm_rare <- repClonality(.immdata_filter$data, .method = "rare")
  rare.p <- vis(.imm_rare, .meta = .immdata_filter$meta) +
            vis(.imm_rare, .by = "Genotype", .meta = .immdata_filter$meta)
  ggsave2(filename = "rare_clonal_proportion_by_Genotype.pdf", 
          plot = rare.p, path = .plots_dir, height = 8, width = 11)

  # Analyse relative abundance (also known as clonal space homeostasis), which is defined as the proportion of repertoire occupied by clonal groups with specific abundances:

  .imm_hom <- repClonality(.immdata_filter$data,
                          .method = "homeo",
                          .clone.types = c(Small = .0001, Medium = .001, 
                                           Large = .01, Hyperexpanded = 1)
                          )
  
  hom.p <- vis(.imm_hom, .meta = .immdata_filter$meta) + 
           vis(imm_hom, .by = c("Genotype"), .meta = .immdata_filter$meta)
  ggsave2(filename = "rel_abundance_clonotypes.pdf", 
          plot = hom.p, path = .plots_dir, height = 8, width = 11)
  
  # Build the heatmap of public clonotypes shared between repertoires
  
  overlap.p <- repOverlap(.immdata_filter$data) %>% vis()  
  ggsave2(filename = "clonotype_overlap.pdf", 
          plot = overlap.p, path = .plots_dir, height = 8, width = 11)

  
  # # Calculate the Chao1 diversity of repertoires, grouped by genotype

    diversity.p <- repDiversity(.immdata_filter$data) %>% vis(.by = c("Genotype"), .meta = .immdata_filter$meta)  
  ggsave2(filename = "clonotype_diversity.pdf", 
          plot = diversity.p, path = .plots_dir, height = 7, width = 7)

}
```

Run immunarch analysis on each cluster
```{r}
for (c_num in seurat_clusters){
  print(paste0("Processing cluster ",c_num))
  run_immunarch(imput_dir = paste0("./TCR_BY_CLUSTER/CLUSTER_", c_num))
}
```

Test
```{r}

tcr_barcodes <- c()

for (s in LN_immdata_filter$meta$Sample){
  # Create vector of barcodes from Barrcode column
  bc <- strsplit(x = paste(LN_immdata_filter$data[[s]]$Barcode, 
                           collapse = ";"), split = ";")[[1]]
  
  # add sample name
  bc = str_replace(string = paste(s,bc, sep = '_'), 
              pattern = "TCR_", 
              replacement = "")
  
  tcr_barcodes <- c(tcr_barcodes, bc)
}
length(tcr_barcodes)

# Filter seurat object keeping only barcodes in TCR dataset
seurat_tcr <- ln_merged_seurat_filtered
seurat_tcr@meta.data$sample = rownames(seurat_tcr@meta.data)
seurat_tcr <- subset(seurat_tcr, subset = sample  %in% tcr_barcodes)

```

Seurat analysis on cells with TCR data
```{r}
annot_clusters.p1 <- DimPlot(seurat_tcr, reduction = "umap", label = TRUE, repel = TRUE, group.by = 'genotype') + ylim(-15,11) # UMAP for TCR cells 
annot_clusters.p2 <- DimPlot(ln_merged_seurat_filtered, reduction = "umap", label = TRUE, repel = TRUE, group.by = 'customclassif') + ylim(-15,11) # UMAP for ALL cells 
ggsave2(filename = "./Plots/TCR_ln_annotated_clusters_umap.pdf", plot = annot_clusters.p1 + annot_clusters.p2, width = 21, height = 11)
annot_clusters.p1

annot_clusters.p3 <- DimPlot(seurat_tcr, reduction = "umap", label = TRUE, repel = TRUE, split.by = 'genotype', group.by = 'customclassif') + ylim(-15,11)
ggsave2(filename = "./Plots/TCR_ln_annotated_clusters_umap_split.pdf", 
        plot = annot_clusters.p3, width = 21, height = 11)
```

### Estimate proportion of each cell type per sample
```{r}
ln_meta_tcr <- as_tibble(seurat_tcr@meta.data)

# Remove unwanted columns
keep <- grep(pattern = '^RNA_snn_res', colnames(ln_meta_tcr), invert = T)
ln_meta_tcr <- ln_meta_tcr[,keep]
```

#### Add columns with data summaries grouped by genotype, mouse ID and cell-type
```{r}
cell_dist_tcr <- ln_meta_tcr %>% group_by(genotype, mouse) %>%
    count(genotype, customclassif) %>%
    mutate(percent = n / sum(n) * 100) %>%
    mutate(group = paste0(genotype,"_",customclassif))

colnames(cell_dist_tcr) <- c("Genotype","Mouse_ID","Cell_type","Cell_count","Percent","Group" )

write.csv(x = cell_dist_tcr, file = "Cell_distribution_summary_TCR.csv",col.names = F )
head(cell_dist_tcr)
```

#### Cell-type composition plots
```{r}
cell_dist.p <- cell_dist_tcr %>% ggplot(aes(fill=Cell_type, y=Cell_count, x=Genotype)) +
    geom_bar(position="fill", stat="identity") +
    scale_fill_viridis(discrete = T) +
    ggtitle("Proportion of cell-types per genotype") +
    theme_ipsum(base_family = 'Helvetica', base_size = 10, axis_title_just = 0.5) +
    ylab('Percent') 

ggsave2(filename = "./Plots/cell_distrib_barchar_TCR.pdf", plot = cell_dist.p)
print(cell_dist.p)
```

```{r}
cell_dist_perc_tcr.p <- cell_dist_tcr %>% ggplot(aes(fill=Cell_type, y=Percent, x=Genotype)) +
                    geom_bar(position="dodge", stat="identity") +
                    scale_fill_viridis(discrete = T) +
                    ggtitle("Proportion of cell-types per genotype (TCR)") +
                    theme_ipsum(axis_title_just = 0.5, base_family = 'Helvetica', strip_text_size=10) +
                    facet_wrap(~Cell_type, labeller = label_wrap_gen())


ggsave2(filename = "./Plots/cell_distrib_barchar_facet_TCR.pdf", plot = cell_dist_perc_tcr.p, width = 13, height = 13)
print(cell_dist_perc_tcr.p)

```
```{r}
cell_dist.box_tcr.p <- cell_dist_tcr %>% ggplot(aes(x = Genotype, y = Percent, fill=Cell_type)) + 
                                  geom_boxplot() + 
                                  scale_fill_viridis(discrete = T) +
                                  theme_ipsum(axis_title_just = 0.5, 
                                              base_family = 'Helvetica', 
                                              strip_text_size=10) +
                                  facet_wrap(~Cell_type, labeller = label_wrap_gen())

ggsave2(filename = "./Plots/cell_distrib_boxplot_facet_tcr.pdf", plot = cell_dist.box_tcr.p, width = 13, height = 13)
print(cell_dist.box_tcr.p)
```
```{r}
sessionInfo()
```
